The state learner -- a super learner for right-censored data

### 2025-06-06 manuscript/manuscript-one.tex
incorporate tag changes

Accurately predicting risk from time-to-event data is a central
challenge in diverse research fields including epidemiology, economics
and weather forecasting with applications in clinical
decision making and policy interventions. For instance, in prostate
cancer management, clinicians often need to estimate a patient’s risk
of disease progression and mortality over time to make informed
decisions about treatment strategies such as active surveillance
versus immediate intervention. Reliable time-to-event risk predictions
can help tailor care to individual patients, avoid overtreatment, and
allocate healthcare resources more effectively. Super learning
\citep{van2007super}, also known as ensemble learning or stacked
regression \citep{wolpert1992stacked,breiman1996stacked}, provides a
powerful approach to this problem which by combining multiple
candidate prediction models reduces the risk of bias incurred by a
single potentially mispecified model. In survival analysis, a super
learner may combine a stack of Cox regression models with a stack of
random survival forests \citep[][Section 8.4]{gerds2021medical}. Such
a strategy has recently produced KDpredict
(\url{https://kdpredict.com/}) a model which jointly predicts the
risks of kidney failure and all-cause mortality at multiple time
horizons based on different sets of covariates
\citep{liu2024predicting}. To evaluate the prediction performance of
the learners the super learner behind KDpredict uses inverse
probability of censoring weighting (IPCW) where the censoring
distribution is estimated under the restrictive assumption that the
censoring distribution does not depend on the covariates. This is a
potential source of bias which is difficult to overcome with the
currently available methods.

In this paper we propose the state learner, a new super learner
designed to handle the specific challenges of ensemble learning with
right-censored data. The state learner simultaneously learns
prediction models for the event-time and censoring distributions. The
state learner is based on the simple idea of an artificial competing
risk model where censoring is included as a state of its
own. Candidate learners of the event time hazard and censoring hazard
functions are then assessed based on how well they predict the state
occupation probabilities of the artificial competing risk model across
time based on baseline covariate information. Our estimation framework
allows for competing risks, avoids restrictive assumptions on the
censoring distribution, and is also fully flexible with respect to the
choice of learners. The latter is in contrast to other proposals which
restrict the library of learners to specific model classes
\citep{polley2011-sl-cens,golmakani2020super}, see Section
\ref{sec:choice-of-loss-function}. Recent work by
\cite{han2021inverse} and \cite{westling2021inference} also
circumvents the need to pre-specify a censoring model by iterating
between estimation of the outcome and censoring models. However, this
procedure is in general not guaranteed to converge to the true
data-generating mechanism \citep[][Appendix~A.4]{munch2024thesis}. To
analyse the theoretical properties of the state learner we focus on
the discrete super learner which picks the model in the library with
best estimated performance \citep{van2007super}. Here we provide
theoretical guarantees for the performance of the state learner and in
particular show that the discrete state learner is consistent under
natural conditions and prove a finite-sample oracle inequality. We
demonstrate how to construct a library for the state learner using
common survival models, and how to obtain risk predictions from the
resulting ensemble.\marginnote{is the term cross-fitting correct?}
The state learner can be applied to produce a risk
prediction model and can also be used for cross-fitting in semiparametric inference.

\subsection{Outline of the paper}
\label{sec:outline-paper}

We introduce our notation and framework in
Section~\ref{sec:framework}. Section~\ref{sec:super-learning}
introduces loss-based super learning. In
Section~\ref{sec:super-learner-simple} we define the state learner,
while Section~\ref{sec:theor-results-prop} provides theoretical
guarantees. In Section~\ref{sec:use-cases-state} we demonstrate how
the state learner can be used to obtain risk predictions, and we
briefly mention how it can be applied in the context of causal
inference.  Section~\ref{sec:numer-exper} reports the results of
numerical experiments, and Section~\ref{sec:real-data-appl}
illustrates the method on prostate cancer data. We conclude with a
discussion in Section~\ref{sec:discussion}. Proofs are collected in
Appendix~\ref{sec:proofs}. Code and an implementation of the state
learner are available at \url{https://github.com/amnudn/statelearner}.

@@

The main idea of the state learner is to jointly use learners for \(
\Lambda_1 \), \( \Lambda_2 \), and \( \Gamma \), and the relations in
equation~(\ref{eq:parametrizeP}), to learn a feature of the observed
data distribution \( P \). The discrete state learner ranks a tuple of
learners for the tuple of the cumulative hazard functions \(
(\Lambda_1, \Lambda_2, \Gamma) \) based on how well they jointly model
the observed data. To formally introduce the state learner, we define
the process 
\begin{equation*}
  \eta(t) = \1\{\tilde{T} \leq t, \tilde D=1\} + 2\,\1\{\tilde{T} \leq t, \tilde
  D=2\} - \1\{\tilde{T} \leq t, \tilde D=0\},
  \quad \text{for} \quad t \in [0, \tau]
\end{equation*} which takes values in \( \{-1,0,1,2\}\) which
represent four mutually exclusive states. Specifically value \( 0 \)
represents the state where the individual is still event-free and
uncensored, value \( 1\) the state where the event of interest has
occured, value \( 2\) the state where a competing risk has occured and
value \( -1\) the state where the observation is right censored.  The
state occupation probabilities given baseline covariates \( X \) are
given by the function\marginnote{cannot use k because k is used in
previous section}
\begin{equation}
  \label{eq:F-def}
  F(t, k, x) = P(\eta(t) = k \mid X=x),
\end{equation}
for all \( t \in [0,\tau] \), \( k \in \{-1,0,1,2\} \), and
\( x \in \mathcal{X} \).

The state learner is a super learner for the function-valued parameter
$\theta(P) = F$ which is identified through
equation~(\ref{eq:F-def}). Under conditional independent censoring
each tuple $(\Lambda_{1}, \Lambda_{2}, \Gamma, H)$ characterises a
distribution \(P\in\mathcal P\), c.f.\
equation~\eqref{eq:parametrizeP}, which in turn determines \( (F, H)
\). Hence, a learner for \( F \) can be constructed from learners for
\( \Lambda_1 \), \( \Lambda_2 \), and $\Gamma$ as follows:
\begin{equation}\label{eq:transition}
  \begin{split}
    F(t, 0, x)
    &
      = P(\tilde{T} > t \mid X= x)
      =
      \exp{{\{-\Lambda_{1}(t \mid x)-\Lambda_{2}(t \mid x) - \Gamma(t \mid x)\}
      }},
    \\
    F(t, 1, x)
    &
      = P(\tilde{T} \leq t, \tilde{D}=1 \mid X=x)
      = \int_0^t F(s-, 0, x)  \Lambda_{1}(\diff s \mid x),
    \\
    F(t, 2, x)
    &
      = P(\tilde{T} \leq t, \tilde{D}=2 \mid X=x)
      = \int_0^t  F(s-, 0, x)  \Lambda_{2}(\diff s \mid x),
    \\
    F(t, -1, x)
    &
      = P(\tilde{T} \leq t, \tilde{D}=0 \mid X=x)
      = \int_0^t F(s-, 0, x)  \Gamma(\diff s \mid x).
  \end{split}
\end{equation}
Equation~\eqref{eq:transition} implies that a library for the
state learner can by build from three libraries of learners:
\(\mathcal{A}_1\), \( \mathcal{A}_2 \), and \( \mathcal{B} \),
where \(\mathcal{A}_1\) and \( \mathcal{A}_2\) contain
learners for the conditional cause-specific cumulative hazard
functions \(\Lambda_1\) and \( \Lambda_2\), respectively, and
\(\mathcal{B}\) contains learners for the conditional
cumulative hazard function of the censoring distribution.
Based on the Cartesian product of libraries of learners for
\((\Lambda_1,\Lambda_2,\Gamma)\) we construct a library
$\mathcal{F}$ of learners for \( F \):
\begin{align*}
  \mathcal{F}(\mathcal{A}_1, \mathcal{A}_2, \mathcal{B})
  &= \{ \phi_{a_1,a_2, b} : a_1 \in \mathcal{A}_1, a_2 \in \mathcal{A}_2, b \in \mathcal{B}\},
    \intertext{where in correspondence with  the relations in equation
    \eqref{eq:transition},}
    \phi_{a_1,a_2, b}(\data_n)(t,0,x)
  &= \exp{\{-a_1(\data_n)(s \mid x)-a_2(\data_n)(s \mid x) - b(\data_n)(s \mid
    x)\} },
  \\
  \phi_{a_1,a_2, b}(\data_n)(t,1,x)
  &= \int_0^t
    \phi_{a_1,a_2, b}(\data_n)(s-,0,x)  a_1(\data_n)(\diff s \mid x),
  \\
  \phi_{a_1,a_2, b}(\data_n)(t,2,x)
  &= \int_0^t\phi_{a_1,a_2, b}(\data_n)(s-,0,x)  a_2(\data_n)(\diff s \mid x),
  \\
  \phi_{a_1,a_2, b}(\data_n)(t,-1,x)
  &= \int_0^t \phi_{a_1,a_2, b}(\data_n)(s-,0,x)  b(\data_n)(\diff s \mid x).
\end{align*} Notably, the libraries \( \mathcal{A}_1 \), \(
\mathcal{A}_2 \), and \( \mathcal{B} \) can be constructed using
common methods for survival analysis.\marginnote{Elaborate on this last statement with a concrete example?}

To evaluate how well a function \( F \) predicts the
process $\eta$ we use the integrated Brier score \citep{graf1999assessment}
\( \bar B_\tau( F,O) = \int_0^{\tau} B_t(F,O) \diff t \),
where \( B_t \) is the Brier score
\citep{brier1950verification} at time \( t \in [0, \tau] \),
\begin{equation*}
  B_t(F,O) = \sum_{j=-1}^{2}
  \left(
      F(t,j,X) - \1{\{\eta(t)=j\}}
  \right)^2.
\end{equation*}
\marginnote{Say that Brier score is across all 4 states because this is not commonly done}
Based on a split of a data set \(\data_n\) into $K$ disjoint
approximately equally sized subsets (c.f., Section
\ref{sec:super-learning}), each learner
\( \phi_{a_1, a_2, b} \) in the library
\( \mathcal{F}(\mathcal{A}_1, \mathcal{A}_2, \mathcal{B}) \)
is evaluated using the cross-validated loss,
\begin{equation*}
  \hat{R}_{n}(\phi_{a_1,a_2,b} ; \bar{B}_{\tau}) =
  \frac{1}{K}\sum_{k=1}^{K}
  \frac{1}{| \data_n^{k} |}\sum_{O_i \in \data_n^{k}}
  \bar B_\tau
  {
    \left(
      \phi_{a_1,a_2,b}{ (\data_n^{-k})}
      , O_i
    \right)
  },
\end{equation*}
and the discrete state learner is given by
\begin{align*}\label{eq:discrete-state-learner}
  \hat{\phi}_n
  &=  \argmin_{(a_1,a_2,b)\in \mathcal{A}_1\times\mathcal{A}_2\times\mathcal{B}}
    \hat{R}_{n}(\phi_{a_1,a_2,b} ; \bar{B}_{\tau}).
\end{align*}

@@

\begin{proposition}
  \label{prop:stric-prop}
  If \(P \in\mathcal{P}\) then
  \begin{equation*}
    F_P = \argmin_{F \in \mathcal{H}_{\mathcal{P}}} P{[\bar{B}_\tau(F, \blank)]},
  \end{equation*}
  for all \( j \in \{-1, 0, 1, 2 \} \), almost all
  \( t \in [0,\tau] \), and \( P \)-almost all
  \( x \in \mathcal{X} \).
\end{proposition}
\begin{proof}
  See Appendix~\ref{sec:consistency-proof}.
\end{proof}

@@

To establish a finite sample oracle result for the state learner we
assume a split of the data into equally sized folds, and for
simplicity of presentation we take \( n \) to be such that \(
|\data_n^{-k}| = n/K \) with \( K \) fixed. We will allow the number
of learners to grow with \( n \) and write \(
\mathcal{F}_n=\mathcal{F}(\mathcal{A}_{1,n}, \mathcal{A}_{2,n},
\mathcal{B}_n)\) as short-hand notation emphasing the
dependence on the sample size. In the following we let the space \(
\mathcal{H}_{\mathcal{P}}= \{F_P : P \in \mathcal{P}\} \) be equipped with the norm \( \| \blank
\|_{P} \) defined by
\begin{equation}
  \label{eq:norm}
  \| F \|_{P} = 
  \left\{
    \sum_{j=-1}^{2}\int_{\mathcal{X}} \int_0^{\tau} F(t, j, x)^2 \diff t P( \diff x)
  \right\}^{1/2}.
\end{equation}

\begin{proposition}
  \label{prop:oracle-prop}
  For all \(P\in\mathcal{P}\), \( n \in \N \), \( k \in \{1, \dots, K\} \),
  and $\delta>0$,
  \begin{align*}
    \E_{P}{\left[ \Vert \hat{\phi}_n(\data_n^{-k}) - F_P \Vert_{P}^2 \right]}
    & \leq (1 + 2\delta)
      \E_{P}{\left[ \Vert \tilde{\phi}_n(\data_n^{-k}) - F_P \Vert_{P}^2 \right]}
    \\
    & \quad
      + (1+ \delta) 16   K \tau
      \left(
      13 + \frac{12}{\delta}
      \right)
      \frac{\log(1 + |\mathcal{F}_n|)}{n}.
  \end{align*}
\end{proposition}
\begin{proof}
  See Appendix~\ref{sec:oracle-inequalities}.
\end{proof}
\marginnote{if it is a special case, why is a proof needed?}
Proposition~\ref{prop:oracle-prop} is in essence a
special case of Theorem 2.3 in \citep{vaart2006oracle}. It 
has the following asymptotic consequences.

@@

For all super learners we use a library consisting of three
learners: The Kaplan-Meier estimator
\citep{kaplan1958nonparametric,Gerds_2019prodlim}, a Cox model
with main effects \citep{cox1972regression, survival-package},
and a random survival forest
\citep{ishwaran2008random,rfsrc-paclage}. We use the same
library to learn the outcome distribution and the censoring
distribution. Note that the three learners in our library of
learners can be used to learn the cumulative hazard functions
of the outcome and the censoring distribution. The latter
works by training the learner on the data set \( \data_n^c \),
where \( \data_n^c = \{O_i^c\}_{i=1}^n \) with
\( O_i^c = (\tilde{T}_i, 1-\Delta_i, X_i) \). When we say that
we use a learner for the cumulative hazard function of the
outcome to learn the cumulative hazard function of the
censoring time, we mean that the learner is trained on
\( \data_n^c \).

@@

\begin{lemma}
  \label{lemma:norm}
  \( R_{P}(F) = \Vert F - F_P \Vert_{P}^2 \).
\end{lemma}
\begin{proof}[Proof]
  For any \( t \in [0, \tau] \) and \( j\in \{-1,0,1,2\} \) we have
  \begin{align*}
    & \E_{P}{\left[ (F(t, j, X) - \1{\{\eta(t) = j \}})^2 \right]}
    \\
    & =    \E_{P}{\left[ (F(t, j, X) - F_P(t, j, X) + F_P(t, j, X) - \1{\{\eta(t) = j
      \}})^2 \right]}
    \\
    & =    \E_{P}{\left[ (F(t, j, X) - F_P(t, j, X))^2\right]}
      + \E_{P}{\left[ (F_P(t, j, X) - \1{\{\eta(t) = j \}})^2\right]}
    \\
    & \quad
      + 2\E_{P}{\left[ (F(t, j, X) - F_P(t, j, X))(F_P(t, j, X) - \1{\{\eta(t) = j
      \}})\right]}
    \\
    & =    \E_{P}{\left[ (F(t, j, X) - F_P(t, j, X))^2\right]}
      + \E_{P}{\left[ (F_P(t, j, X) - \1{\{\eta(t) = j \}})^2\right]},
  \end{align*}
  where the last equality follows from the tower property. Hence, using Fubini,
  we have
  \begin{equation*}
    P{[\bar{B}_{\tau}(F, \blank)]}
    = \Vert F - F_P \Vert_{P}^2 + P{[\bar{B}_{\tau}(F_P, \blank)]}.
  \end{equation*}
\end{proof}

>>>>

Accurately predicting risk from time-to-event data is a central
challenge in diverse research fields including epidemiology, economics
and weather forecasting with applications in clinical
decision making and policy interventions. For instance, in prostate
cancer management, clinicians often need to estimate a patient’s risk
of disease progression and mortality over time to make informed
decisions about treatment strategies such as active surveillance
versus immediate intervention. Reliable time-to-event risk predictions
can help tailor care to individual patients, avoid overtreatment, and
allocate healthcare resources more effectively. Super learning
\citep{van2007super}, also known as ensemble learning or stacked
regression \citep{wolpert1992stacked,breiman1996stacked}, provides a
powerful approach to this problem by combining multiple
candidate prediction models to reduce the risk of bias incurred by a
single potentially mispecified model. In survival analysis, a super
learner may combine a stack of Cox regression models with a stack of
random survival forests \citep[][Section 8.4]{gerds2021medical}. Such
a strategy has recently produced KDpredict
(\url{https://kdpredict.com/}) a model which jointly predicts the
risks of kidney failure and all-cause mortality at multiple time
horizons based on different sets of covariates
\citep{liu2024predicting}. To evaluate the prediction performance of
the learners the super learner behind KDpredict uses inverse
probability of censoring weighting (IPCW) where the censoring
distribution is estimated under the restrictive assumption that the
censoring distribution does not depend on the covariates. This is a
potential source of bias which is difficult to overcome with the
currently available methods.

In this paper we propose the state learner, a new super learner
designed to handle the specific challenges of ensemble learning with
right-censored data. The state learner simultaneously learns
prediction models for the event-time and censoring distributions. The
state learner is based on the simple idea of an artificial competing
risk model where censoring is included as a state of its
own. Candidate learners of the event time hazard and censoring hazard
functions are then assessed based on how well they predict the state
occupation probabilities of the artificial competing risk model across
time based on baseline covariate information. Our estimation framework
allows for competing risks, avoids restrictive assumptions on the
censoring distribution, and is also fully flexible with respect to the
choice of learners. The latter is in contrast to other proposals which
restrict the library of learners to specific model classes
\citep{polley2011-sl-cens,golmakani2020super}, see Section
\ref{sec:choice-of-loss-function}. Like our proposal, recent work by
\cite{han2021inverse} and \cite{westling2021inference} also
circumvents the need to pre-specify a censoring model by iterating
between estimation of the outcome and censoring models.
However, this iterative
procedure is in general not guaranteed to converge to the true
data-generating mechanism
\citep[][Appendix~A.4]{munch2024thesis}.

To analyse the theoretical properties of the state learner we
focus on the discrete super learner which picks the model in
the library with best estimated performance
\citep{van2007super}. We provide theoretical guarantees
for the performance of the state learner and in particular
show that the discrete state learner is consistent under
natural conditions and prove a finite-sample oracle
inequality. We demonstrate how to construct a library for the
state learner using common survival models, and how to obtain
risk predictions from the resulting ensemble.

The rest of the paper is organized as follows.
We introduce our notation and framework in
Section~\ref{sec:framework}. Section~\ref{sec:super-learning}
introduces loss-based super learning. In
Section~\ref{sec:super-learner-simple} we define the state learner,
while Section~\ref{sec:theor-results-prop} provides theoretical
guarantees. In Section~\ref{sec:use-cases-state} we demonstrate how
the state learner can be used to obtain risk predictions, and we
briefly mention how it can be applied in the context of causal
inference.  Section~\ref{sec:numer-exper} reports the results of
numerical experiments, and Section~\ref{sec:real-data-appl}
illustrates the method on prostate cancer data. We conclude with a
discussion in Section~\ref{sec:discussion}. Proofs are collected in
Appendix~\ref{sec:proofs}. Code and an implementation of the state
learner are available at \url{https://github.com/amnudn/statelearner}.

@@

The main idea of the state learner is to jointly use learners for \(
\Lambda_1 \), \( \Lambda_2 \), and \( \Gamma \), and the relations in
equation~(\ref{eq:parametrizeP}), to learn a feature of the observed
data distribution \( P \). The discrete state learner ranks a tuple of
learners for the tuple of the cumulative hazard functions \(
(\Lambda_1, \Lambda_2, \Gamma) \) based on how well they jointly model
the observed data. To formally introduce the state learner, we define
the process 
\begin{equation*}
  \eta(t) = \1\{\tilde{T} \leq t, \tilde D=1\} + 2\,\1\{\tilde{T} \leq t, \tilde
  D=2\} - \1\{\tilde{T} \leq t, \tilde D=0\},
  \quad \text{for} \quad t \in [0, \tau],
\end{equation*}
which takes values in \( \{-1,0,1,2\}\). The four values
represent four mutually exclusive states. Specifically, value
\( 0 \) represents the state where the individual is still
event-free and uncensored, value \( 1\) the state where the
event of interest has occurred, value \( 2\) the state where a
competing risk has occurred, and value \( -1\) the state where
the observation is right-censored. The state occupation
probabilities given baseline covariates \( X \) are given by
the function
\begin{equation}
  \label{eq:F-def}
  F(t, l, x) = P(\eta(t) = l \mid X=x),
\end{equation}
for all \( t \in [0,\tau] \), \( l \in \{-1,0,1,2\} \), and
\( x \in \mathcal{X} \).

The state learner is a super learner for the function-valued parameter
$\theta(P) = F$ which is identified through
equation~(\ref{eq:F-def}). Under conditional independent censoring
each tuple $(\Lambda_{1}, \Lambda_{2}, \Gamma, H)$ characterises a
distribution \(P\in\mathcal P\), c.f.\
equation~\eqref{eq:parametrizeP}, which in turn determines \( (F, H)
\). Hence, a learner for \( F \) can be constructed from learners for
\( \Lambda_1 \), \( \Lambda_2 \), and $\Gamma$ as follows:
\begin{equation}\label{eq:transition}
  \begin{split}
    F(t, 0, x)
    &
      = P(\tilde{T} > t \mid X= x)
      =
      \exp{{\{-\Lambda_{1}(t \mid x)-\Lambda_{2}(t \mid x) - \Gamma(t \mid x)\}
      }},
    \\
    F(t, 1, x)
    &
      = P(\tilde{T} \leq t, \tilde{D}=1 \mid X=x)
      = \int_0^t F(s-, 0, x)  \Lambda_{1}(\diff s \mid x),
    \\
    F(t, 2, x)
    &
      = P(\tilde{T} \leq t, \tilde{D}=2 \mid X=x)
      = \int_0^t  F(s-, 0, x)  \Lambda_{2}(\diff s \mid x),
    \\
    F(t, -1, x)
    &
      = P(\tilde{T} \leq t, \tilde{D}=0 \mid X=x)
      = \int_0^t F(s-, 0, x)  \Gamma(\diff s \mid x).
  \end{split}
\end{equation}
Equation~\eqref{eq:transition} implies that a library for the
state learner can by build from three libraries of learners:
\(\mathcal{A}_1\), \( \mathcal{A}_2 \), and \( \mathcal{B} \),
where \(\mathcal{A}_1\) and \( \mathcal{A}_2\) contain
learners for the conditional cause-specific cumulative hazard
functions \(\Lambda_1\) and \( \Lambda_2\), respectively, and
\(\mathcal{B}\) contains learners for the conditional
cumulative hazard function of the censoring distribution.
Based on the Cartesian product of libraries of learners for
\((\Lambda_1,\Lambda_2,\Gamma)\) we construct a library
$\mathcal{F}$ of learners for \( F \):
\begin{align*}
  \mathcal{F}(\mathcal{A}_1, \mathcal{A}_2, \mathcal{B})
  &= \{ \phi_{a_1,a_2, b} : a_1 \in \mathcal{A}_1, a_2 \in \mathcal{A}_2, b \in \mathcal{B}\},
    \intertext{where in correspondence with  the relations in equation
    \eqref{eq:transition},}
    \phi_{a_1,a_2, b}(\data_n)(t,0,x)
  &= \exp{\{-a_1(\data_n)(s \mid x)-a_2(\data_n)(s \mid x) - b(\data_n)(s \mid
    x)\} },
  \\
  \phi_{a_1,a_2, b}(\data_n)(t,1,x)
  &= \int_0^t
    \phi_{a_1,a_2, b}(\data_n)(s-,0,x)  a_1(\data_n)(\diff s \mid x),
  \\
  \phi_{a_1,a_2, b}(\data_n)(t,2,x)
  &= \int_0^t\phi_{a_1,a_2, b}(\data_n)(s-,0,x)  a_2(\data_n)(\diff s \mid x),
  \\
  \phi_{a_1,a_2, b}(\data_n)(t,-1,x)
  &= \int_0^t \phi_{a_1,a_2, b}(\data_n)(s-,0,x)  b(\data_n)(\diff s \mid x).
\end{align*} Notably, the libraries \( \mathcal{A}_1 \), \(
\mathcal{A}_2 \), and \( \mathcal{B} \) can be constructed using
standard software for survival analysis. In
\texttt{R}, for instance, we can construct Cox models as
learners using the \texttt{survival}-package
\citep{survival-package}, and we can construct random survival
forests as learners using the \texttt{randomForestSRC}-package \citep{randomForestSRC}.

To evaluate how well a function \( F \) predicts the
process $\eta$ we use the integrated Brier score \citep{graf1999assessment}
\( \bar B_\tau( F,O) = \int_0^{\tau} B_t(F,O) \diff t \),
where \( B_t \) is the Brier score
\citep{brier1950verification} at time \( t \in [0, \tau] \),
\begin{equation*}
  B_t(F,O) = \sum_{l=-1}^{2}
  \left(
      F(t,l,X) - \1{\{\eta(t)=l\}}
  \right)^2.
\end{equation*}
\marginnote{Say that Brier score is across all 4 states because this is not commonly done}
Based on a split of a data set \(\data_n\) into $K$ disjoint
approximately equally sized subsets (c.f., Section
\ref{sec:super-learning}), each learner
\( \phi_{a_1, a_2, b} \) in the library
\( \mathcal{F}(\mathcal{A}_1, \mathcal{A}_2, \mathcal{B}) \)
is evaluated using the cross-validated loss,
\begin{equation*}
  \hat{R}_{n}(\phi_{a_1,a_2,b} ; \bar{B}_{\tau}) =
  \frac{1}{K}\sum_{k=1}^{K}
  \frac{1}{| \data_n^{k} |}\sum_{O_i \in \data_n^{k}}
  \bar B_\tau
  {
    \left(
      \phi_{a_1,a_2,b}{ (\data_n^{-k})}
      , O_i
    \right)
  },
\end{equation*}
and the discrete state learner is given by
\begin{align*}\label{eq:discrete-state-learner}
  \hat{\phi}_n
  &=  \argmin_{(a_1,a_2,b)\in \mathcal{A}_1\times\mathcal{A}_2\times\mathcal{B}}
    \hat{R}_{n}(\phi_{a_1,a_2,b} ; \bar{B}_{\tau}).
\end{align*}

@@

\begin{proposition}
  \label{prop:stric-prop}
  If \(P \in\mathcal{P}\) then
  \begin{equation*}
    F_P = \argmin_{F \in \mathcal{H}_{\mathcal{P}}} P{[\bar{B}_\tau(F, \blank)]},
  \end{equation*}
  for all \( l \in \{-1, 0, 1, 2 \} \), almost all
  \( t \in [0,\tau] \), and \( P \)-almost all
  \( x \in \mathcal{X} \).
\end{proposition}
\begin{proof}
  See Appendix~\ref{sec:consistency-proof}.
\end{proof}

@@

To establish a finite sample oracle result for the state learner we
assume a split of the data into equally sized folds, and for
simplicity of presentation we take \( n \) to be such that \(
|\data_n^{-k}| = n/K \) with \( K \) fixed. We will allow the number
of learners to grow with \( n \) and write \(
\mathcal{F}_n=\mathcal{F}(\mathcal{A}_{1,n}, \mathcal{A}_{2,n},
\mathcal{B}_n)\) as short-hand notation emphasing the
dependence on the sample size. In the following we let the space \(
\mathcal{H}_{\mathcal{P}}= \{F_P : P \in \mathcal{P}\} \) be equipped with the norm \( \| \blank
\|_{P} \) defined by
\begin{equation}
  \label{eq:norm}
  \| F \|_{P} = 
  \left\{
    \sum_{l=-1}^{2}\int_{\mathcal{X}} \int_0^{\tau} F(t, l, x)^2 \diff t P( \diff x)
  \right\}^{1/2}.
\end{equation}

\begin{proposition}
  \label{prop:oracle-prop}
  For all \(P\in\mathcal{P}\), \( n \in \N \), \( k \in \{1, \dots, K\} \),
  and $\delta>0$,
  \begin{align*}
    \E_{P}{\left[ \Vert \hat{\phi}_n(\data_n^{-k}) - F_P \Vert_{P}^2 \right]}
    & \leq (1 + 2\delta)
      \E_{P}{\left[ \Vert \tilde{\phi}_n(\data_n^{-k}) - F_P \Vert_{P}^2 \right]}
    \\
    & \quad
      + (1+ \delta) 16   K \tau
      \left(
      13 + \frac{12}{\delta}
      \right)
      \frac{\log(1 + |\mathcal{F}_n|)}{n}.
  \end{align*}
\end{proposition}
\begin{proof}
  See Appendix~\ref{sec:oracle-inequalities}.
\end{proof}

Proposition~\ref{prop:oracle-prop} has the following
asymptotic consequences.

@@

For all super learners we use a library consisting of three
learners: The Kaplan-Meier estimator
\citep{kaplan1958nonparametric,Gerds_2019prodlim}, a Cox model
with main effects \citep{cox1972regression, survival-package},
and a random survival forest
\citep{ishwaran2008random,randomForestSRC}. We use the same
library to learn the outcome distribution and the censoring
distribution. Note that the three learners in our library of
learners can be used to learn the cumulative hazard functions
of the outcome and the censoring distribution. The latter
works by training the learner on the data set \( \data_n^c \),
where \( \data_n^c = \{O_i^c\}_{i=1}^n \) with
\( O_i^c = (\tilde{T}_i, 1-\Delta_i, X_i) \). When we say that
we use a learner for the cumulative hazard function of the
outcome to learn the cumulative hazard function of the
censoring time, we mean that the learner is trained on
\( \data_n^c \).

@@

\begin{lemma}
  \label{lemma:norm}
  \( R_{P}(F) = \Vert F - F_P \Vert_{P}^2 \).
\end{lemma}
\begin{proof}[Proof]
  For any \( t \in [0, \tau] \) and \( l\in \{-1,0,1,2\} \) we have
  \begin{align*}
    & \E_{P}{\left[ (F(t, l, X) - \1{\{\eta(t) = l \}})^2 \right]}
    \\
    & =    \E_{P}{\left[ (F(t, l, X) - F_P(t, l, X) + F_P(t, l, X) - \1{\{\eta(t) = l
      \}})^2 \right]}
    \\
    & =    \E_{P}{\left[ (F(t, l, X) - F_P(t, l, X))^2\right]}
      + \E_{P}{\left[ (F_P(t, l, X) - \1{\{\eta(t) = l \}})^2\right]}
    \\
    & \quad
      + 2\E_{P}{\left[ (F(t, l, X) - F_P(t, l, X))(F_P(t, l, X) - \1{\{\eta(t) = l
      \}})\right]}
    \\
    & =    \E_{P}{\left[ (F(t, l, X) - F_P(t, l, X))^2\right]}
      + \E_{P}{\left[ (F_P(t, l, X) - \1{\{\eta(t) = l \}})^2\right]},
  \end{align*}
  where the last equality follows from the tower property. Hence, using Fubini,
  we have
  \begin{equation*}
    P{[\bar{B}_{\tau}(F, \blank)]}
    = \Vert F - F_P \Vert_{P}^2 + P{[\bar{B}_{\tau}(F_P, \blank)]}.
  \end{equation*}
\end{proof}
